\bf Looking at Non-Linear Dimension Reductions as Models in the Data Space

### 2024-09-06 paper.qmd
Initial writing for the model

The MSE for different bin widths ($a_1$) using tSNE, UMAP, PHATE, PaCMAP, and TriMAP with default (hyper-)parameter setting (@fig-mnist-mse) were calculated. It is found that tSNE (@fig-mnist-mse (a)) provide the most reasonable representation for the digit 1 dataset, showing universally best. However, \gD{} representation shows a big non-linear cluster and a small cluster with a small gap.

In the case of digit 1 data, there should not be any clusters unless anomalies exist, indicating different digit 1 patterns. The angle of the digit 1 images varies along this non-linear clustering structure (@fig-tsne-best), while the small cluster contains the digit 1 images with different patterns of the digit 1, unlike the usual (@fig-model-error-mnist (c)).

@@

```{r}
#| label: fig-model-error-mnist
#| fig-cap: "$10\\text{-}D$ model error in tSNE layout of MNIST digit 1 dataset in $2\\text{-}D$. Most low $10\\text{-}D$ model errors are distributed along the lower edge of the $2\\text{-}D$ structure, while most large $10\\text{-}D$ model errors are concentrated along the upper edge (a). There are certain data points that exhibit large error rates due to their deviation from the usual $10\\text{-}D$ data structure, which makes them anomalies. These anomalies can be classified into two types: those that are anomalies within the non-linear structure and those that lie outside of it. The images associated with high model error points within the non-linear structure display different patterns of the digit 1 (b). However, when comparing these images to the ones found outside of the non-linear structure, it becomes evident that the latter display different patterns of the digit 1 (c). "
#| fig-pos: H
#| fig-width: 4
#| fig-height: 8
#| out-height: 80%

>>>>

The MSE for different bin widths ($a_1$) using tSNE, UMAP, PHATE, PaCMAP, and TriMAP with default (hyper-)parameter setting (@fig-mnist-mse) were calculated. It is found that tSNE (@fig-mnist-mse (a)) provide the most reasonable representation for the digit 1 dataset, showing universally best. However, \gD{} representation shows a big non-linear cluster and a small cluster with a small gap (@fig-tsne-best).

In the case of digit 1 data, there should not be any clusters unless anomalies exist, indicating different digit 1 patterns. The angle of the digit 1 images varies along this non-linear clustering structure (@fig-tsne-best), while the small cluster contains the digit 1 images with different patterns of the digit 1, unlike the usual (@fig-model-error-mnist (c)). This provides the evidence for two close clusters.

The *model-in-the-data-space* helps to assess the \gD{} layout. As shown in @mnist-model-proj (a), the model provides the evidence for the non-linear structure in $10\text{-}D$. Also, by visualizing the model in $10\text{-}D$ shows some quirks. The model has a twisted pattern which actually provide evidence for the $10\text{-}D$ structure which couldn't be observed by \gD{} layout (@@mnist-model-proj (b)). Furthermore, there are some long edges which indicates the existence of the small cluster and it's so close (@@mnist-model-proj (c)).

@@

```{r}
#| label: fig-model-error-mnist
#| fig-cap: "(a) The $10\\text{-}D$ model error for tSNE layout in $\\gD{}$ of MNIST digit 1 dataset. Most low $10\\text{-}D$ model errors are distributed along the big non-linear clutser, while most large $10\\text{-}D$ model errors are concentrated along the small cluster. (b) There are certain data points that exhibit large error rates due to their deviation from the usual $10\\text{-}D$ data structure, which makes them anomalies. These anomalies can be classified into two types: those that are anomalies within the non-linear structure and those that lie outside of it. The images associated with high model error points within the non-linear structure display different patterns of the digit 1. However, when comparing these images to the ones found outside of the non-linear structure, it becomes evident that the latter display different patterns of the digit 1 (c). "
#| fig-pos: H
#| fig-width: 4
#| fig-height: 8
#| out-height: 80%

### 2024-09-06 paper.tex
Initial writing for the model

The MSE for different bin widths (\(a_1\)) using tSNE, UMAP, PHATE,
PaCMAP, and TriMAP with default (hyper-)parameter setting
(Figure~\ref{fig-mnist-mse}) were calculated. It is found that tSNE
(Figure~\ref{fig-mnist-mse} (a)) provide the most reasonable
representation for the digit 1 dataset, showing universally best.
However, \gD{} representation shows a big non-linear cluster and a small
cluster with a small gap.

In the case of digit 1 data, there should not be any clusters unless
anomalies exist, indicating different digit 1 patterns. The angle of the
digit 1 images varies along this non-linear clustering structure
(Figure~\ref{fig-tsne-best}), while the small cluster contains the digit
1 images with different patterns of the digit 1, unlike the usual
(Figure~\ref{fig-model-error-mnist} (c)).

@@

\caption{\label{fig-model-error-mnist}\(10\text{-}D\) model error in
tSNE layout of MNIST digit 1 dataset in \(2\text{-}D\). Most low
\(10\text{-}D\) model errors are distributed along the lower edge of the
\(2\text{-}D\) structure, while most large \(10\text{-}D\) model errors
are concentrated along the upper edge (a). There are certain data points
that exhibit large error rates due to their deviation from the usual
\(10\text{-}D\) data structure, which makes them anomalies. These
anomalies can be classified into two types: those that are anomalies
within the non-linear structure and those that lie outside of it. The
images associated with high model error points within the non-linear
structure display different patterns of the digit 1 (b). However, when
comparing these images to the ones found outside of the non-linear
structure, it becomes evident that the latter display different patterns
of the digit 1 (c).}

>>>>

The MSE for different bin widths (\(a_1\)) using tSNE, UMAP, PHATE,
PaCMAP, and TriMAP with default (hyper-)parameter setting
(Figure~\ref{fig-mnist-mse}) were calculated. It is found that tSNE
(Figure~\ref{fig-mnist-mse} (a)) provide the most reasonable
representation for the digit 1 dataset, showing universally best.
However, \gD{} representation shows a big non-linear cluster and a small
cluster with a small gap (Figure~\ref{fig-tsne-best}).

In the case of digit 1 data, there should not be any clusters unless
anomalies exist, indicating different digit 1 patterns. The angle of the
digit 1 images varies along this non-linear clustering structure
(Figure~\ref{fig-tsne-best}), while the small cluster contains the digit
1 images with different patterns of the digit 1, unlike the usual
(Figure~\ref{fig-model-error-mnist} (c)). This provides the evidence for
two close clusters.

The \emph{model-in-the-data-space} helps to assess the \gD{} layout. As
shown in \citet{mnist-model-proj} (a), the model provides the evidence
for the non-linear structure in \(10\text{-}D\). Also, by visualizing
the model in \(10\text{-}D\) shows some quirks. The model has a twisted
pattern which actually provide evidence for the \(10\text{-}D\)
structure which couldn't be observed by \gD{} layout
(@\citet{mnist-model-proj} (b)). Furthermore, there are some long edges
which indicates the existence of the small cluster and it's so close
(@\citet{mnist-model-proj} (c)).

@@

\caption{\label{fig-model-error-mnist}(a) The \(10\text{-}D\) model
error for tSNE layout in \(\gD{}\) of MNIST digit 1 dataset. Most low
\(10\text{-}D\) model errors are distributed along the big non-linear
clutser, while most large \(10\text{-}D\) model errors are concentrated
along the small cluster. (b) There are certain data points that exhibit
large error rates due to their deviation from the usual \(10\text{-}D\)
data structure, which makes them anomalies. These anomalies can be
classified into two types: those that are anomalies within the
non-linear structure and those that lie outside of it. The images
associated with high model error points within the non-linear structure
display different patterns of the digit 1. However, when comparing these
images to the ones found outside of the non-linear structure, it becomes
evident that the latter display different patterns of the digit 1 (c).}
